---
title: "Lipidomics Analysis by Functional Group"
author: "Emily Bean"
date: "March 12, 2020"
output: 
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```
This ANOVA structure is for the functional groups of lipids instead of each individual annotated lipid (example - Sphingosines).

# Pairwise comparisons  
*All comparisons made for both tumor and plasma tissues*

1. Plasma vs tumor
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance 


```{r}

# read in data
require(tidyverse)
require(ggplot2)
require(emmeans)
require(data.table)

# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read all lipids data
# this is a big dataset and will be slow
lip <- fread("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allLipidsCleaned.txt", header = TRUE)  %>% 
  # add column of functional groups
  mutate(group = as.factor(sapply(str_split(lipid, " "), `[`, 1)),
         # set log.area to numberic
         log.area = as.numeric(log.area)) %>% 
  # for now: remove w/o
  filter(!group == "w/o")

# make separate tumor and plasma dataframes for easier analysis 
plasma <- lip %>% 
  filter(!tissue.type == "tumor") %>% 
  mutate(Time = factor(str_extract(SampleName, "(\\d{1,2})"), 
                       ordered = TRUE, levels = c("7", "21", "35")))
# tumor df 
tumor <- filter(lip, tissue.type == "tumor")


```


```{r}

## define function to perform an ANOVA, post-hoc emmeans, and return significant data

myAnova <- function(df, variable, lipidGroups) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(lipidGroups)) {
    
    # create function
    f <- as.formula(paste0("log.area ~", variable))
    
    # perform ANOVA
    mod <- aov(f, data = filter(df, group == lipidGroups[i]))
    
    # collect variables
    pvals <- data.frame(lipid = lipidGroups[i],
                        fStat =  round(summary(mod)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e <- as.data.frame(emmeans(mod, f1)$contrasts,
                       rownames = FALSE)
    
    # append to outDF
    out <- cbind(pvals, e)
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}

```


```{r}

# define a function to perform the ANOVA and post-hoc for each of the three Times


myAnovaTime <- function(variable, df, lipidGroups) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(lipidGroups)) {
    
    # create function
    f <- as.formula(paste0("log.area ~", variable))
    
    # perform ANOVA
    mod1 <- aov(f, data = filter(df, group == lipidGroups[i] & Time == "7"))
    
    # collect variables
    pvals1 <- data.frame(lipid = lipidGroups[i],
                         Time = "D7",
                         fStat =  round(summary(mod1)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e1 <- as.data.frame(emmeans(mod1, f1)$contrasts,
                        rownames = FALSE)
    
    ## ANOVA Day 21
    # perform ANOVA
    mod2 <- aov(f, data = filter(df, group == lipidGroups[i] & Time == "21"))
    
    # collect variables
    pvals2 <- data.frame(lipid = lipidGroups[i],
                         Time = "D21",
                         fStat =  round(summary(mod2)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e2 <- as.data.frame(emmeans(mod2, f1)$contrasts,
                        rownames = FALSE)
    
    ## ANOVA Day 35
    # perform ANOVA
    mod3 <- aov(f, data = filter(df, group == lipidGroups[i] & Time == "35"))
    
    # collect variables
    pvals3 <- data.frame(lipid = lipidGroups[i],
                         Time = "D35",
                         fStat =  round(summary(mod3)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e3 <- as.data.frame(emmeans(mod3, f1)$contrasts,
                        rownames = FALSE)
    
    # append to outDF
    out <- rbind(cbind(pvals1, e1),
                 cbind(pvals2, e2),
                 cbind(pvals3, e3))
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}


```

## Plasma v tumor

```{r}
# plasma v tumor

mod <- myAnova(df = lip, variable = "tissue.type", lipidGroups = unique(lip$group))

knitr::kable(mod)

```

Below is a table of the significant lipid groups with means and standard deviations.
```{r}

sigsdf <- lip %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(tissue.type, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(tissue.type, stat, sep = "_"),
         tissue.type = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf, col.names = c("lipid", "Plasma mean", 
                                   "Plasma SD", "Tumor mean", "Tumor SD"))

```


## Plasma D7 v D21 v D35


```{r}
# plasma over time

mod <- myAnova(df = plasma, variable = "Time", lipidGroups = unique(lip$group))

knitr::kable(mod)


```


```{r}

sigsdf <- plasma %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(Time, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(Time, stat, sep = "_"),
         Time = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```

## Tumor

### 4 Treatment groups

```{r}
# plasma v tumor

mod <- myAnova(df = tumor, variable = "treatmentID", lipidGroups = unique(tumor$group))

knitr::kable(mod)

```

Below is a table of the significant lipid groups with means and standard deviations.
```{r}

sigsdf <- tumor %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(treatmentID, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(treatmentID, stat, sep = "_"),
         treatmentID = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```

### Exercise vs Sedentary

```{r}
# plasma v tumor

mod <- myAnova(df = tumor, variable = "Exercise", lipidGroups = unique(tumor$group))

knitr::kable(mod)

```

Below is a table of the significant lipid groups with means and standard deviations.
```{r}

sigsdf <- tumor %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(Exercise, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(Exercise, stat, sep = "_"),
         Exercise = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```

### Weight gain vs Weight maintenance


```{r}
# plasma v tumor

mod <- myAnova(df = tumor, variable = "Weight", lipidGroups = unique(tumor$group))

knitr::kable(mod)

```

Below is a table of the significant lipid groups with means and standard deviations.
```{r}

sigsdf <- tumor %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(Weight, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(Weight, stat, sep = "_"),
         Weight = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```

## Plasma

### 4 Treatment groups


```{r}

# ANOVA
mod <- myAnovaTime(variable = "treatmentID", df = plasma, lipidGroups = unique(plasma$group))
  
knitr::kable(mod)


```

```{r}

sigsdf <- plasma %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(treatmentID, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(treatmentID, stat, sep = "_"),
         treatmentID = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```

### Exercise vs Sedentary

```{r}

# ANOVA
mod <- myAnovaTime(variable = "Exercise", df = plasma, lipidGroups = unique(plasma$group))
  
knitr::kable(mod)


```

```{r}

sigsdf <- plasma %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(Exercise, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(Exercise, stat, sep = "_"),
         Exercise = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```


### Weight gain vs Weight maintenance

```{r}

# ANOVA
mod <- myAnovaTime(variable = "Weight", df = plasma, lipidGroups = unique(plasma$group))
  
knitr::kable(mod)


```

```{r}

sigsdf <- plasma %>% 
  # get only significant lipids
  filter(group %in% unique(as.character(mod$lipid))) %>% 
  group_by(Weight, group) %>% 
  # get the mean and sd
  summarize(mean = round(mean(log.area, na.rm = TRUE), 3),
            sd = round(sd(log.area, na.rm = TRUE), 3)) %>% 
  # gather and spread to get correct layout
  gather(key = stat, value = value, mean, sd) %>% 
  ungroup() %>% 
  mutate(t.s = paste(Weight, stat, sep = "_"),
         Weight = NULL,
         stat = NULL) %>% 
  spread(key = t.s, value = value)

knitr::kable(sigsdf)

```