---
title: "Lipidomics PCA"
author: "Emily Bean"
date: "2/12/2020"
output:
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```

# Overview

This script performs a principal components analysis (PCA) on each data subset and then creates several diagnostics plots, then a grouping PCA and a biplot. 

The top eigenvalues are shown for each PCA.


# Pairwise comparisons  
*All comparisons made for both tumor and plasma tissues*

1. Plasma vs tumor
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance  

```{r}
## get data

require(dplyr)
require(tidyr)
require(factoextra)
require(FactoMineR)
require(corrplot)

# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read custom PCA function
source("https://raw.githubusercontent.com/EmilyB17/ganda-lab/master/MyPCAFunction.R")

# read all lipids data
lip <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allLipidsCleaned.txt", header = TRUE, stringsAsFactors = FALSE) 

## Make all data horizontal for PCAs
liph <- lip %>% 
  select(-ion) %>% 
  spread(key = lipid, value = log.area)

# plasma only
plasmah <- liph %>% 
  filter(tissue.type == "plasma") %>% 
  mutate(Time = sapply(strsplit(SampleName, "_"), `[`, 1)) %>% 
  mutate(Time = factor(Time, ordered = TRUE, levels = c("D7", "D21", "D35")))

# tumor only
tumorh <- liph %>% 
  filter(tissue.type == "tumor")

```





## Plasma vs tumor

```{r}

# perform PCA
data <- liph %>% select(-c(SampleName, tissue.type, mouseID, treatmentID, Exercise, Weight))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = liph$tissue.type,
         groupname = "Tissue Type")
```


## Plasma D7 vs D21 vd D35 

```{r}

# perform PCA
data <- plasmah %>% select(-c(SampleName, tissue.type, mouseID, treatmentID, Exercise, Weight, Time))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping =  plasmah$Time,
         groupname = "Plasma - Time")
```

## Tumor 

### 4 treatment groups (2x2 factorial)

```{r}

# perform PCA
data <- tumorh %>% select(-c(SampleName, tissue.type, mouseID, treatmentID, Exercise, Weight))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$treatmentID,
         groupname = "Tumor - 2x2 factorial")
```

### Exercise vs sedentary


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$Exercise,
         groupname = "Tumor - Exercise")
```

### Weight gain vs weight maintenance


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$Weight,
         groupname = "Tumor - Weight")
```

## Plasma

### 4 treatment groups

```{r}

# perform PCA
data <- plasmah %>% select(-c(SampleName, tissue.type, mouseID, treatmentID, Exercise, Weight, Time))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$treatmentID,
         groupname = "Plasma - 2x2 factorial")
```

### Exercise vs sedentary


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$Exercise,
         groupname = "Plasma - Exercise")
```

### Weight gain vs weight maintenance


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$Weight,
         groupname = "Plasma - Weight")
```



