---
title: "Preliminary Metabolomics Analysis"
author: "Emily Bean"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

## Overview

This is the preliminary data analysis for the 15-02 metabolomics dataset, Dr. Connie Rogers lab. 

**General cleaning steps:**  
1. Download archived raw .wiff files from Metabolomics Core  
2. Convert .wiff to .mzML in ProteoWizard 4.0 (MS Convert)  
3. Peak picking and alignment in MSDIAL (MoNA database for hydrophilic; LipidBlast for lipids)  
4. Tidy in Excel  
5. Upload to MetaboAnalyst; quantile normalization, Pareto scaling, and log transformation  
6. Download normalized data for statistical analysis in R  

**Analysis Steps:**  
1. Individual metabolites 
2. Sample clustering    
3. Metabolic "community"    

**Pairwise comparisons:**  
*All comparisons made for aqueous (hydrophilic) and lipid data*  
1. Plasma vs. tumor  
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance  



```{r, warning = FALSE, message = FALSE, collapse = TRUE}
## Read in all data and load packages
require(dplyr)
require(tidyr)
require(vegan)
require(ggplot2)
require(pairwiseAdonis)
require(ggordiplots)

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE)

# read all lipids data
lip <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allLipidsCleaned.txt", header = TRUE, stringsAsFactors = TRUE)
```

*Summary of the data*  
There are `r length(unique(aq$metabolite))` unique metabolites in the aqueous dataset. There are `r length(unique(lip$lipid)) ` unique lipids in the lipid dataset. The aqueous metabolites are named, for example: 
```{r warning = FALSE, echo = FALSE}
# table of several example aqueous metabolites
knitr::kable(unique(aq$metabolite)[1:10], col.names = NULL)
```

The lipid metabolites are named based on the organic chemistry and bond order. For example: 
```{r warning = FALSE, echo = FALSE}
# table of several example aqueous metabolites
knitr::kable(unique(lip$lipid)[1:10], col.names = NULL)
```
**Note:** The ` w/o` refers to lipids that had a good match with MS1 but not MS2. These are still usable but we are not as confident in their annotation as we are with the other lipids.  
**Another Note:** These are _annotated metabolites only_. In the raw dataset are many tens of thousands more metabolites and lipids that were detected as compounds but not annotated; we can do a community analysis on these later if we want to.  

## Plasma aqueous metabolites through time  

It's important for downstream analysis to first determine if the plasma metabolome change significantly from day 7, to day 21, to day 35. If there are differences, we will need to keep the days separate for downstream analyses; if not, we can likely lump them all together.  

An ANOVA on each individual metabolite by Time shows that there are several metabolites that differ between the times, so when analyzing plasma metabolome we will keep Time as a separate factor.

```{r}
# create a column for the different days

plasmaaq <- aq %>% 
  filter(!Label == "tumor") %>% 
  mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))

# are there differences between time points?
plasmetab <- unique(plasmaaq$metabolite)
pvals <- data.frame()

for(i in 1:length(plasmetab)) {
  mod <- aov(area ~ Time, data = filter(plasmaaq, metabolite == plasmetab[i]))
  pvals[i, "metabolite"] <- plasmetab[i]
  pvals[i, "pval"] <- round(summary(mod)[[1]][["Pr(>F)"]][1], 4)
}

# pull significant p values into a table
sigs <- pvals[pvals$pval < 0.05,]
knitr::kable(sigs)
```

We know that individual metabolites vary throughout time, and we can also look at the entire metabolite "communitY" with an NMDS. The statistical test is a permutational ANOVA, which can be done in a pair-wise fashion to get p values similar to a Tukey's test.
```{r}

```

The plot does not intuitively show any differences between the ellipses; remember, this is only a visual tool. 
```{r}
# pull NMDS coordinates
plasmaH$NMDS1 <- ord$points[,1]
plasmaH$NMDS2 <- ord$points[,2]

# IF ANY EFT OBJECTS ARE SIGNIFICANT:
# pull significant envfit objects to plots
#species.scores <- data.frame(eft$vectors$arrows[eft$vectors$pvals < 0.05,])
#species.scores$species <- row.names(eft$vectors$arrows[eft$vectors$pvals < 0.05,])

# save gg_ordiplot object to get ellipse values
plot <-  gg_ordiplot(ord, groups = plasmaH$Time, label = FALSE, plot = FALSE)
# get ellipse coordinates
df_ell <- plot$df_ellipse
# get label coordinates for ellipse centers
NMDS.mean <- plot$df_mean.ord
# pull NMDS coordinates
ord.data <- plot$df_ord 

## create in ggplot2
ggplot(data = plasmaH, aes(x = NMDS1, y = NMDS2)) +
  geom_path(data = df_ell, aes(x = x, y = y, color = Group), show.legend = FALSE) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = Time), size = 1) +
  annotate("text",x = NMDS.mean$x, y = NMDS.mean$y,label=NMDS.mean$Group) +
 # geom_segment(data = species.scores,
  #             aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2), colour = "grey") +
  #geom_text(data = species.scores, 
   #         aes(x = NMDS1, y = NMDS2, label=species),
    #        size = 3) +
  ggtitle("Plasma metabolome by Time") +
  theme_bw()
```

## Individual Metabolites 

This is an example of pairwise comparisons for individual aqueous metabolites. The code below runs a one-way ANOVA between the area under the curve (dependent variable) and the four treatment groups (independent variable) for the tumor tissue only. This ANOVA repeats in a "for loop" for all 136 aqueous metabolites, and stores the p values in a table.  
There are four significant metabolites; that is, four metabolites vary significantly between the four treatment groups. 

```{r collapse = TRUE}
### TUMOR

# make separate dataframe for the tumor tissue
tumor <- filter(aq, Label == "tumor")
metabs <- levels(tumor$metabolite)

pvals <- data.frame()

# run the for loop
for(i in 1:length(metabs)) {
  mod <- aov(area ~ treatmentID, data = filter(tumor, metabolite == metabs[i]))
  pvals[i, "metabolite"] <- metabs[i]
  pvals[i, "pval"] <- round(summary(mod)[[1]][["Pr(>F)"]][1], 4)
   
}

# pull significant p values into a table
sigs <- pvals[pvals$pval < 0.05,]
knitr::kable(sigs, col.names =  c("metabolite", "p value"))

```



## Community Analysis

Another way of analyzing this data is to treat all of the metabolites in a sample as a community and look at differences between the "metabolomes" of each sample. 

### Non-metric multidimensional scaling (NMDS)

NMDS plots each sample by its similarity to the community of another sample. The statistical test is a permutational ANOVA to determine statistical differences between treatment groups. In theory, you can visually see statistical significance on the NMDS plot when two ellipses are separate from each other. In practice, this doesn't always happen because ordinations are performed in multiple dimensions and reduced to 2D for visualization, so keep that in mind. 
