---
title: "Metabolomic Analysis"
author: "Emily Bean"
date: "January 10, 2020"
output:
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```

# Overview

This is the preliminary data analysis for the 15-02 metabolomics dataset, Dr. Connie Rogers lab. 


## General cleaning steps 
1. Download archived raw .wiff files from Metabolomics Core  
2. Convert .wiff to .mzML in ProteoWizard 4.0 (MS Convert)  
3. Peak picking and alignment in MSDIAL (MoNA database for hydrophilic; LipidBlast for lipids)  
4. Tidy in Excel  
5. Upload to MetaboAnalyst; quantile normalization, Pareto scaling, and log transformation  
6. Download normalized data for statistical analysis in R  

## Analysis Steps  
1. Individual metabolites (THIS SCRIPT)
2. Sample clustering    
3. Metabolic "community" (`metabolomicsNMDS.Rmd`)   

# Pairwise comparisons  
*All comparisons made for both tumor and plasma tissues*

1. Plasma vs tumor
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance  

**Note:** These are _annotated metabolites only_. In the raw dataset are many tens of thousands more metabolites and lipids that were detected as compounds but not annotated; we can do a community analysis on these later if we want to.

```{r echo = FALSE, include = FALSE}
## Read in all data and load packages
require(dplyr)
require(tidyr)
#require(vegan)
require(ggplot2)
require(emmeans)
#require(pairwiseAdonis)
#require(ggordiplots)

# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

# make separate tumor and plasma dataframes for easier analysis 
plasma <- aq %>% 
  filter(!Label == "tumor") %>% 
  mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))
# tumor df 
tumor <- filter(aq, Label == "tumor")

# get the separate metabolites for each dataframe for downstream analysis
metabs <- unique(aq$metabolite)
plasmetabs <- unique(plasma$metabolite)
tumetabs <- unique(tumor$metabolite)
```


```{r}
## define function to perform an ANOVA, post-hoc emmeans, and return significant data

myAnova <- function(variable, df, metabList) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(metabList)) {
    
    # create function
    f <- as.formula(paste0("area ~", variable))
    
    # perform ANOVA
    mod <- aov(f, data = filter(df, metabolite == metabList[i]))
    
    # collect variables
    pvals <- data.frame(metabolite = metabList[i],
                        fStat =  round(summary(mod)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e <- as.data.frame(emmeans(mod, f1)$contrasts,
                       rownames = FALSE)
    
    # append to outDF
    out <- cbind(pvals, e)
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}
```

```{r}
# define a function to perform the ANOVA and post-hoc for each of the three Times


myAnovaTime <- function(variable, df, metabList) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(metabList)) {
    
    # create function
    f <- as.formula(paste0("area ~", variable))
    
    # perform ANOVA
    mod1 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "7"))
    
    # collect variables
    pvals1 <- data.frame(metabolite = metabList[i],
                         Time = "D7",
                        fStat =  round(summary(mod1)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e1 <- as.data.frame(emmeans(mod1, f1)$contrasts,
                       rownames = FALSE)
    
    ## ANOVA Day 21
    # perform ANOVA
    mod2 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "21"))
    
    # collect variables
    pvals2 <- data.frame(metabolite = metabList[i],
                         Time = "D21",
                         fStat =  round(summary(mod2)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e2 <- as.data.frame(emmeans(mod2, f1)$contrasts,
                        rownames = FALSE)
    
    ## ANOVA Day 35
    # perform ANOVA
    mod3 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "35"))
    
    # collect variables
    pvals3 <- data.frame(metabolite = metabList[i],
                         Time = "D35",
                         fStat =  round(summary(mod3)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e3 <- as.data.frame(emmeans(mod3, f1)$contrasts,
                        rownames = FALSE)
    
    # append to outDF
    out <- rbind(cbind(pvals1, e1),
                  cbind(pvals2, e2),
                  cbind(pvals3, e3))
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}

```

**Note: in pairwise comparions, the data has been normalized, scaled, and log-transformed in MetaboAnalyst. An accompanying PDF report describes those steps.**

## Plasma vs tumor

We would assume that the plasma and tumor tissue would differ significantly in their metabolic makeup, so this is really more of a "double-check" than an an analysis.



```{r}
# is there a difference between tissues?

tissues <- myAnova(variable = "tissue.type",
                   df = aq,
                   metabList = metabs)

knitr::kable(tissues)
```


```{r}
## create table of means

tukdf <- aq[aq$metabolite %in% tissues$metabolite, ]

meansdf <- tukdf %>% 
  group_by(tissue.type, metabolite) %>% 
  summarize(mean = round(mean(area), 3)) %>% 
  spread(key = tissue.type, value = mean)

knitr::kable(meansdf, col.names = c("metabolite", "Plasma mean", "Tumor mean"))


```


```{r}
## make boxplots to illustrate differences

tukdf <- aq[aq$metabolite %in% tissues$metabolite, ]

ggplot(data = tukdf, aes(x = tissue.type, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) 

```

Since there are only two explanatory variables (plasma and tumor), this is technically a t-test so no Tukey's post-hoc is performed. Boxplots visulize the differences and a table of means is shown. Interestingly, the plasma metabolites are all higher than the tumor metabolites. There seem to be several high outliers, even in the normalized data.The boxplots also illustrate that most metabolites have a positive log change in the plasma tissue and a negative log change in the tumor tissue.

## Plasma D7 vs D21 v D35

It's important for downstream analysis to first determine if the plasma metabolome change significantly from day 7, to day 21, to day 35. If there are differences, we will need to keep the days separate for downstream analyses; if not, we can likely lump them all together.  

```{r}


# are there differences between time points?

time <- myAnova(variable = "Time",
                   df = plasma,
                   metabList = plasmetabs)

knitr::kable(time)

```

Another visualization tool is boxplots.

```{r}

## make boxplots to illustrate differences

plotdf <- plasma[plasma$metabolite %in% time$metabolite, ]

# split into three sets of plots so it's easier to see 

plotdf1 <- plotdf[plotdf$metabolite %in% unique(plotdf$metabolite)[1:8],]
plotdf2 <- plotdf[plotdf$metabolite %in% unique(plotdf$metabolite)[9:16],]
plotdf3 <- plotdf[plotdf$metabolite %in% unique(plotdf$metabolite)[17:23],]

# plot 1
ggplot(data = plotdf1, aes(x = Time, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) 
# plot 2
ggplot(data = plotdf2, aes(x = Time, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) 
# plot 3
ggplot(data = plotdf3, aes(x = Time, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) 

```


An ANOVA on each individual metabolite by Time shows that there are multiple metabolites that differ between the times, so when analyzing plasma metabolome we will keep Time as a separate factor. The Tukey tests indicate that most metabolites differ between either D7 and D35, or D21 and D35. Few (or none) indicate difference between D7 and D21.
This should be explored further to examine if the differences are due to cancer development or if they can be attributed to other things like growth and metabolism.

## Four treatment groups

The initial analyis is comparing all four treatment groups to each other in the 2x2 factorial style. The code follows this structure for pairwise comparisons with ANOVA:

>This code is a one-way ANOVA between the log-transformed area under the curve (dependent variable) and the four treatment groups (independent variable) for the tumor tissue only. This ANOVA repeats in a "for loop" for all 136 aqueous metabolites, and stores the p values in a table. 

### Tumor tissue 

There are four significant metabolites; that is, four metabolites vary significantly between the four treatment groups in the tumor tissue. 

```{r collapse = TRUE}
### TUMOR

# is there a difference between treatment types in tumor?

sigs <- myAnova(variable = "treatmentID",
                   df = tumor,
                   metabList = tumetabs)

knitr::kable(sigs)

```



```{r}

## make boxplots to illustrate differences

tukdf <- tumor[tumor$metabolite %in% sigs$metabolite, ]

ggplot(data = tukdf, aes(x = treatmentID, y = area)) +
  geom_boxplot() +
  facet_wrap(~metabolite)
  
```

Quinolinate and Glucose both vary between two different treatments, which is why they are listed twice. It is important to note that dividing the data into four treatment groups rather than two (e.g. Exercise vs Sedentary) decreases the number of mice in each group, which is likely why there are fewer metabolites that are significant when comparing all 4 groups.

### Plasma


Since we clearly saw that there is a difference in the plasma aqueous metabolites between days (above), we will keep Time as a separate factor moving forward. Therefore, there are 3 separate ANOVAs performed for each test; one for each time point. The results of each time point are independent of the others.

```{r}

# is there a difference between treatments in Plasma?

sigs <- myAnovaTime(variable = "treatmentID",
                    df = plasma,
                    metabList = plasmetabs)

knitr::kable(sigs)
              
```



```{r}

## make boxplots to illustrate differences

plotdf <- plasma[plasma$metabolite %in% sigs$metabolite, ]

# split into three sets of plots so it's easier to see 

plotdf1 <- plotdf[plotdf$Time %in% "7",]
plotdf2 <- plotdf[plotdf$Time %in% "21",]
plotdf3 <- plotdf[plotdf$Time %in% "35",]

# plot 1
ggplot(data = plotdf1, aes(x = treatmentID, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) +
  ggtitle("Day 7")
# plot 2
ggplot(data = plotdf2, aes(x = treatmentID, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite)  +
  ggtitle("Day 21")
# plot 3
ggplot(data = plotdf3, aes(x = treatmentID, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) +
  ggtitle("Day 35")

```


## Exercise vs. Sedentary

Two exercise groups: exercised and exercise-restricted (sedentary).

### Tumor

There is one significant metabolite: Quinolinate.

```{r collapse = TRUE}
### TUMOR

# Difference between Exercise treatments in tumor tissue

sigs <- myAnova(variable = "Exercise",
                df = tumor,
                metabList = tumetabs)

knitr::kable(sigs)

```



```{r}
## make boxplots to illustrate differences

tukdf <- tumor[tumor$metabolite %in% sigs$metabolite, ]

ggplot(data = tukdf, aes(x = Exercise, y = area)) +
  geom_boxplot() +
  facet_wrap(~metabolite) 

```


Quinolinate is higher in the Sedentary than Exercising group.

### Plasma

```{r}

# Difference between Exercise treatments in tumor tissue

sigs <- myAnovaTime(variable = "Exercise",
                df = plasma,
                metabList = plasmetabs)

knitr::kable(sigs)

```

```{r}
## make the table of means

tukdf <- plasma[plasma$metabolite %in% sigs$metabolite, ]

meansdf <- tukdf %>% 
  group_by(Exercise, metabolite) %>% 
  summarize(mean = round(mean(area), 3)) %>% 
  spread(key = Exercise, value = mean)

knitr::kable(meansdf, col.names = c("metabolite", "EX mean", "SED mean"))

```

```{r}

## make boxplots to illustrate differences
tukdf <- plasma[plasma$metabolite %in% sigs$metabolite, ]

ggplot(data = tukdf, aes(x = Exercise, y = area)) +
  geom_boxplot() +
  facet_wrap(~metabolite) 

```


There are multiple significant metabolites between Exercise treatments in the plasma tissue, so these are difficult to visualize with boxplots. 


## Weight Restricted vs Ad-lib Fed

Two weight groups: energy restricted and ad libitum fed

### Tumor


```{r}

### TUMOR

# Difference in tumor tissue between Weight treatments

sigs <- myAnova(variable = "Weight",
                df = tumor,
                metabList = tumetabs)

knitr::kable(sigs)

```

```{r}
## make the table of means

tukdf <- tumor[tumor$metabolite %in% sigs$metabolite, ]

meansdf <- tukdf %>% 
  group_by(Weight, metabolite) %>% 
  summarize(mean = round(mean(area), 3)) %>% 
  spread(key = Weight, value = mean)

knitr::kable(meansdf, col.names = c("metabolite", "AL mean", "ER mean"))

```

```{r}
## make boxplots to illustrate differences
tukdf <- tumor[tumor$metabolite %in% sigs$metabolite, ]

ggplot(data = tukdf, aes(x = Weight, y = area)) +
  geom_boxplot() +
  facet_wrap(~metabolite) 

```

### Plasma

```{r}

### PLASMA

# Difference in plasma tissue between Weight treatments

sigs <- myAnovaTime(variable = "Weight",
                df = plasma,
                metabList = plasmetabs)

knitr::kable(sigs)

```

```{r}
## make the table of means

tukdf <- plasma[plasma$metabolite %in% sigs$metabolite, ]

meansdf <- tukdf %>% 
  group_by(Weight, metabolite) %>% 
  summarize(mean = round(mean(area), 3)) %>% 
  spread(key = Weight, value = mean)

knitr::kable(meansdf, col.names = c("metabolite", "AL mean", "ER mean"))

```

```{r}

## make boxplots to illustrate differences

plotdf <- plasma[plasma$metabolite %in% sigs$metabolite, ]

# split into three sets of plots so it's easier to see 

plotdf1 <- plotdf[plotdf$Time %in% "7",]
plotdf2 <- plotdf[plotdf$Time %in% "21",]
plotdf3 <- plotdf[plotdf$Time %in% "35",]

# plot 1
ggplot(data = plotdf1, aes(x = Weight, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) +
  ggtitle("Day 7")
# plot 2
ggplot(data = plotdf2, aes(x = Weight, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite)  +
  ggtitle("Day 21")
# plot 3
ggplot(data = plotdf3, aes(x = Weight, y = area)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_boxplot() +
  facet_wrap(~metabolite) +
  ggtitle("Day 35")

```



