---
title: "Metabolomics Two-Way ANOVA"
author: "Emily Bean"
date: "April 14, 2020"
output:
  rmarkdown::github_document
---


This script performs a two-way or three-way ANOVA on the aqueous metabolite fraction. Two-way ANOVA compares Exercise, Weight, and the interaction of both treatments. Three-way ANOVA compares Exercise, Weight, Time, and the interaction of both and all three.

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```


```{r echo = FALSE, include = FALSE}

## Read in all data and load packages
require(tidyverse)
require(ggplot2)
require(emmeans)


# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

# make separate tumor and plasma dataframes for easier analysis 
plasma <- aq %>% 
  filter(!Label == "tumor") %>% 
  mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))
# tumor df 
tumor <- filter(aq, Label == "tumor")


```


```{#r}

# write table of summary statistics

# summary stats on all data (mean, sd)

sum <- aq %>% 
  group_by(tissue.type, metabolite, Exercise, Weight) %>% 
  summarize(mean = mean(area),
            sd = sd(area)) %>% 
  ungroup() %>% 
  mutate(metabolite = str_remove(gsub("\\.", "-", metabolite), "X"))

write.table(sum, file = "C:/Users/emily/OneDrive - The Pennsylvania State University/Research/git/mice-metab/results_tables/aqueous-summary-stats.txt",
            sep = "\t", row.names = FALSE)

```


```{r}

# define a function to perform two-way ANOVA
myTwoWayAnova <- function(df, metabList) {
  
  
  # make output dataframes
  allpvals <- data.frame()
  posthoc_sigs <- data.frame()
  
  # loop through each metabolite
  for(i in 1:length(metabList)) {
    
    # perform ANOVA on the iterated metabolite
    mod <- aov(area ~ Exercise*Weight, data = filter(df, metabolite == metabList[i]))
    
    # collect p values and F statistics
    pvals <- data.frame(metabolite = metabList[i],
                        fStatEx =  round(summary(mod)[[1]][["F value"]][1], 3),
                        pvalEx = round(summary(mod)[[1]][["Pr(>F)"]][1], 3),
                        fStatWt = round(summary(mod)[[1]][["F value"]][2], 3),
                        pvalWt = round(summary(mod)[[1]][["Pr(>F)"]][2], 3),
                        fStatInt = round(summary(mod)[[1]][["F value"]][3], 3),
                        pvalInt = round(summary(mod)[[1]][["Pr(>F)"]][3], 3))
    
    # save to an outDF
    allpvals <- rbind(allpvals, pvals)
    
    # perform Tukey post-hoc
    tukey <- TukeyHSD(mod)
    
    # collect p vals
    posthoc <- as.data.frame(rbind(as.data.frame(tukey[1]$Exercise) %>% 
                                 mutate(comparison = "Exercise",
                                          contrast = rownames(tukey[1]$Exercise)),
                               as.data.frame(tukey[2]$Weight) %>% 
                                 mutate(comparison = "Weight",
                                        contrast = rownames(tukey[2]$Weight)),
                               as.data.frame(tukey[3]$`Exercise:Weight`) %>% 
                                 mutate(comparison = "Interaction",
                                        contrast = rownames(tukey[3]$`Exercise:Weight`))))

    
    # if there are any significant p vals, add to outDF
    if(any(posthoc$`p adj` < 0.05)) {
      
      out <- posthoc %>% filter(`p adj` < 0.05)
      out$metabolite <- metabList[i]
      
      posthoc_sigs <- rbind(posthoc_sigs, out)
      
    }
    
    
  }

  # return the output dataframes
  return(list(allpvals, posthoc_sigs))
}

```


```{r}

# define function to perform three-way ANOVA with Time
myThreeWayAnova <- function(df, metabList) {
  
  
  # make output dataframes
  allpvals <- data.frame()
  posthoc_sigs <- data.frame()
  
  # loop through each metabolite
  for(i in 1:length(metabList)) {
    
    # perform ANOVA on the iterated metabolite
    mod <- aov(area ~ Exercise*Weight*Time, data = filter(df, metabolite == metabList[i]))
    
    # collect p values and F statistics
    pvals <- data.frame(metabolite = metabList[i],
                        fStatEx =  round(summary(mod)[[1]][["F value"]][1], 3),
                        pvalEx = round(summary(mod)[[1]][["Pr(>F)"]][1], 3),
                        fStatWt = round(summary(mod)[[1]][["F value"]][2], 3),
                        pvalWt = round(summary(mod)[[1]][["Pr(>F)"]][2], 3),
                        fStatTime = round(summary(mod)[[1]][["F value"]][3], 3),
                        pvalTime = round(summary(mod)[[1]][["Pr(>F)"]][3], 3),
                        fStatEx_Wt = round(summary(mod)[[1]][["F value"]][4], 3),
                        pvalEx_Wt = round(summary(mod)[[1]][["Pr(>F)"]][4], 3),
                        fStatEx_Time = round(summary(mod)[[1]][["F value"]][5], 3),
                        pvalEx_Time = round(summary(mod)[[1]][["Pr(>F)"]][5], 3),
                        fStatWt_Time = round(summary(mod)[[1]][["F value"]][6], 3),
                        pvalWt_Time = round(summary(mod)[[1]][["Pr(>F)"]][6], 3),
                        fStatEx_Wt_Time = round(summary(mod)[[1]][["F value"]][7], 3),
                        pvalEx_Wt_Time = round(summary(mod)[[1]][["Pr(>F)"]][7], 3))
    
    # save to an outDF
    allpvals <- rbind(allpvals, pvals)
    
    # perform Tukey post-hoc
    tukey <- TukeyHSD(mod)
    
    # collect p vals
    posthoc <- as.data.frame(rbind(as.data.frame(tukey[1]$Exercise) %>% 
                                 mutate(comparison = "Exercise",
                                          contrast = rownames(tukey[1]$Exercise)),
                                   as.data.frame(tukey[2]$Weight) %>% 
                                 mutate(comparison = "Weight",
                                          contrast = rownames(tukey[2]$Weight)),
                                   as.data.frame(tukey[3]$Time) %>% 
                                 mutate(comparison = "Time",
                                          contrast = rownames(tukey[3]$Time)),
                                   as.data.frame(tukey[4]$`Exercise:Weight`) %>% 
                                 mutate(comparison = "Ex-Wt",
                                          contrast = rownames(tukey[4]$`Exercise:Weight`)),
                                   as.data.frame(tukey[5]$`Exercise:Time`) %>% 
                                 mutate(comparison = "Ex-Time",
                                          contrast = rownames(tukey[5]$`Exercise:Time`)),
                                   as.data.frame(tukey[6]$`Weight:Time`) %>% 
                                 mutate(comparison = "Wt-Time",
                                          contrast = rownames(tukey[6]$`Weight:Time`)),
                                   as.data.frame(tukey[7]$`Exercise:Weight:Time`) %>% 
                                 mutate(comparison = "Ex-Wt-Time",
                                          contrast = rownames(tukey[7]$`Exercise:Weight:Time`))))
    
    # if there are any significant p vals, add to outDF
    if(any(posthoc$`p adj` < 0.05)) {
      
      out <- posthoc %>% filter(`p adj` < 0.05)
      out$metabolite <- metabList[i]
      
      posthoc_sigs <- rbind(posthoc_sigs, out)
      
    }
    
    
  }

  # return the output dataframes
  return(list(allpvals, posthoc_sigs))
}

```


## Tumor Tissue; Two-way ANOVA

```{r}

tumormod <- myTwoWayAnova(df = tumor, metabList = unique(tumor$metabolite))

```

**Table of all F statistics & unadjusted p values**

```{#r}

# print table of all pvalues
#knitr::kable(as.data.frame(tumormod[1]))

# wrangle data for nicer table format
# make table of results

tab <- as.data.frame(tumormod[1]) %>% 
  # fix up 'metabolite' name
  mutate(metabolite = str_remove(gsub("\\.", "-", metabolite), "X"))

write.table(tab, "C:/Users/emily/OneDrive - The Pennsylvania State University/Research/git/mice-metab/results_tables/aqueous-twowayANOVA-fstat_pval.txt", sep = "\t",
            row.names = FALSE)


```

**Table of significant and adjusted p values from Tukey's post hoc **

```{r}

# print table of all significant post-hoc findings
sigs <- as.data.frame(tumormod[2]) %>% 
  select(metabolite, comparison, contrast, p.adj) %>% 
  mutate(p.adj = round(p.adj, 3))

#knitr::kable(sigs)


# wrangle data for nicer table format
# make table of results

#tab <- as.data.frame(tumormod[2]) %>% 
  # fix up 'metabolite' name
  mutate(metabolite = str_remove(gsub("\\.", "-", metabolite), "X")) %>% 
  select(metabolite, comparison, contrast, diff, lwr, upr, p.adj)

#write.table(tab, "C:/Users/emily/OneDrive - The Pennsylvania State University/Research/git/mice-metab/results_tables/aqueous-twowayANOVA-oosthoc.txt", sep = "\t",
            row.names = FALSE)

```

**Visualizations**

```{r}

# get metabolites with significant interaction term
sigsInt <- data.frame(tumormod[2]) %>% 
  filter(comparison == "Interaction")

sigdf <- tumor %>% 
  semi_join(sigsInt, by = "metabolite") %>% 
  group_by(metabolite, Exercise, Weight) %>% 
  summarize(mean = mean(area),
            sd = sd(area),
            cilow = mean - 1.96*sd,
            cihi = mean + 1.96*sd) %>% 
  ungroup()

# make a plot
ggplot(data = sigdf, aes(x = Exercise, y = mean, group = Weight, color = Weight)) +
  geom_point() +
  geom_line() +
  geom_errorbar(data = sigdf, aes(ymin = mean + sd, ymax = mean - sd), width = .1) +
  facet_wrap(~metabolite) +
  ggtitle("Significant Interactions") +
  labs(x = "Exercise", y = "Mean log area")

## get metabolites with just Exericse significance
sigEx <- data.frame(tumormod[2]) %>% 
  filter(comparison == "Exercise") 

sigdfEx <- tumor %>% 
  semi_join(sigEx, by = "metabolite")

# plot (just Quinolinate)
ggplot(data = sigdfEx, aes(x = Exercise, y = area)) +
  geom_boxplot() +
  labs(x = "Exercise", y = "Log area", title = "Quinolinate")

## get metabolites with just Weight significance
sigWt <- data.frame(tumormod[2]) %>% 
  filter(comparison == "Weight") 

sigdfWt <- tumor %>% 
  semi_join(sigWt, by = "metabolite") %>% 
  group_by(metabolite, Weight) %>% 
  summarize(mean = mean(area),
            sd = sd(area),
            cilow = mean - 1.96*sd,
            cihi = mean + 1.96*sd) %>% 
  ungroup()

# plot
ggplot(data = sigdfWt, aes(x = Weight, y = mean, group = metabolite, color = metabolite)) +
  geom_point() +
  geom_line() +
  labs(x = "Weight", y = "Mean log area", title = "Significant between Weight treatment")

```


## Plasma; Three-way ANOVA

```{r}

plasmamod <- myThreeWayAnova(df = plasma, metabList = unique(plasma$metabolite))

```

**Table of all F statistics & p values**

```{#r}

# print table of all pvalues
#knitr::kable(as.data.frame(plasmamod[1]))

# wrangle data for nicer table format
# make table of results

tab <- as.data.frame(plasmamod[1]) %>% 
  # fix up 'metabolite' name
  mutate(metabolite = str_remove(gsub("\\.", "-", metabolite), "X"))

write.table(tab, "C:/Users/emily/OneDrive - The Pennsylvania State University/Research/git/mice-metab/results_tables/aqueous-threewayANOVA-fstat_pval.txt", sep = "\t",
            row.names = FALSE)

```


**Table of significant p values from Tukey's post hoc **

```{r}

# print table of all significant post-hoc findings
sigs <- as.data.frame(plasmamod[2]) %>% 
  select(metabolite, comparison, contrast, p.adj) %>% 
  mutate(p.adj = round(p.adj, 3))

#knitr::kable(sigs)

# wrangle data for nicer table format
# make table of results

#tab <- as.data.frame(plasmamod[2]) %>% 
  # fix up 'metabolite' name
  mutate(metabolite = str_remove(gsub("\\.", "-", metabolite), "X")) %>% 
  select(metabolite, comparison, contrast, diff, lwr, upr, p.adj)

#write.table(tab, "C:/Users/emily/OneDrive - The Pennsylvania State University/Research/git/mice-metab/results_tables/aqueous-threewayANOVA-posthoc.txt", sep = "\t",
            row.names = FALSE)

```

**Interactions without Time**

```{r}

# get metabolites with significant interaction term
sigsInt <- data.frame(plasmamod[2]) %>% 
  filter(comparison == "Ex-Wt")

sigdf <- plasma %>% 
  semi_join(sigs, by = "metabolite") %>% 
  group_by(metabolite, Exercise, Weight) %>% 
  summarize(mean = mean(area),
            sd = sd(area),
            cilow = mean - 1.96*sd,
            cihi = mean + 1.96*sd) %>% 
  ungroup()

# make a plot
ggplot(data = sigdf, aes(x = Exercise, y = mean, group = Weight, color = Weight)) +
  geom_point() +
  geom_line() +
  geom_errorbar(data = sigdf, aes(ymin = mean + sd, ymax = mean - sd), width = .1) +
  facet_wrap(~metabolite) +
  ggtitle("Significant Interactions w/o Time") +
  labs(x = "Exercise", y = "Mean log area")

```

**Interactions with Ex, Weight, and Time**

```{r}

# get significant interactions
sigAll <- as.data.frame(plasmamod[2]) %>% 
  filter(comparison == "Ex-Wt-Time")

sigdf <- plasma %>% 
  semi_join(sigAll, by = "metabolite") %>% 
  group_by(metabolite, treatmentID, Time, Exercise, Weight) %>% 
  summarize(mean = mean(area),
            sd = sd(area),
            cilow = mean - 1.96*sd,
            cihi = mean + 1.96*sd) %>% 
  ungroup()

# make a plot
ggplot(data = sigdf, aes(x = Time, y = mean, group = treatmentID, color = treatmentID)) +
  geom_point(aes(shape = Exercise)) +
  geom_line(aes(linetype = Weight)) +
  facet_wrap(~metabolite) +
  ggtitle("Significant Interactions")+
  labs(x = "Exercise", y = "Mean log area")

```

