---
title: "Metabolomic NMDS"
author: "Emily Bean"
date: "January 10, 2020"
output:
  html_document:
    toc: true
    theme: united
---

## Overview

See "metabolomicsAnalysis.Rmd" for details

## Community Analysis

**Note: this data has been processed by MetaboAnalyst but not normalized (see Analysis Report for details on this process)**

```{r echo = FALSE}
# read data
require(ggordiplots)
require(vegan)

# source myNMDS function
source("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/R/FUNCTIONmyNMDS.R")

# read un-normalized data
aqproc <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/aqueousCleanedNotNormalized.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

# make separate tumor and plasma dataframes for easier analysis 
plasmaproc <- aqproc %>% 
  filter(!Label == "tumor") %>% 
  mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))

# this data needs to be horizontal for NMDS
aqproch <- aqproc %>% 
  spread(key = metabolite, value = area) %>% 
  select(-c(id, mouseID))

plasmaproch <- plasmaproc %>% 
  spread(key = metabolite, value = area) %>% 
  select(-c(id, mouseID))

tumorh <- aqproc %>% 
  filter(Label == "tumor") %>% 
    spread(key = metabolite, value = area) %>% 
  select(-c(id, mouseID))


```

#### Plasma vs tumor

As expected, there is a significant difference between tumor and plasma tissue.
```{r}

# plasma vs tumor
abundDF <- aqproch %>% select(-c(Label, treatmentID, Exercise, Weight, tissue.type))
sampDF <- aqproch %>% select(Label, treatmentID, Exercise, Weight, tissue.type)

myNMDS(abundDF, sampDF, indVar = "tissue.type", ordK = 2)

```

#### Plasma D7 vs D21 v D35

There are significant difference between the times.
```{#r}

# plasma days
abundDF <- plasmaproch %>% select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))
sampDF <- plasmaproch %>% select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "Time", ordK = 2)

```

#### Four treatment groups

##### Tumor

None of the four groups were significantly different. 
```{r}
# tumor by treatmentID
abundDF <- tumorh %>% select(-c(Label, treatmentID, Exercise, Weight, tissue.type))
sampDF <- tumorh %>% select(Label, treatmentID, Exercise, Weight, tissue.type)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

##### Plasma

Since there are differences by day, we will do 3 separate ordinations for each time
```{r}

# DAY 7
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "7") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "7") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

```{r}

# DAY 21
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "21") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "21") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

```{#r}

# DAY 35
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "35") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "35") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```


#### Exercise vs Sedentary

##### Tumor


None of the four groups were significantly different. 
```{r}
# tumor by treatmentID
abundDF <- tumorh %>% select(-c(Label, treatmentID, Exercise, Weight, tissue.type))
sampDF <- tumorh %>% select(Label, treatmentID, Exercise, Weight, tissue.type)

myNMDS(abundDF, sampDF, indVar = "Exercise", ordK = 2)
```

##### Plasma

Since there are differences by day, we will do 3 separate ordinations for each time
```{r}

# DAY 7
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "7") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "7") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

```{r}

# DAY 21
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "21") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "21") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

```{#r}

# DAY 35
# plasma by treatmentID
abundDF <- plasmaproch %>% 
  filter(Time == "35") %>% 
  select(-c(Label, treatmentID, Exercise, Weight, tissue.type, Time))

sampDF <- plasmaproch %>% 
  filter(Time == "35") %>% 
  select(Label, treatmentID, Exercise, Weight, tissue.type, Time)

myNMDS(abundDF, sampDF, indVar = "treatmentID", ordK = 2)
```

