---
title: "Metabolomics PermANOVA"
author: "Emily Bean"
date: "February 23, 2020"
output:
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```

# Overview

This script follows the same structure as `metabolmicsAnalysis.Rmd` but performs a permutational ANOVA on the entire metabolomics set. The data is scaled from 0 to 1, and then a Bray-Curtis distance matrix is created. A pairwise permutational ANOVA (PERMANOVA) on the data based on the grouping variable.

This analysis results in a pairwise comparison on the "structure" of the data as a "community".


# Pairwise comparisons  
*All comparisons made for both tumor and plasma tissues*

1. Plasma vs tumor
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance  



```{r}
## get data

require(dplyr)
require(tidyr)
require(vegan)
require(pairwiseAdonis)

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

## Make all data horizontal for PCAs
aqh <- aq %>% 
  spread(key = metabolite, value = area)

# standardize the data
aqhstand <- decostand(aqh[ , 8:ncol(aqh)], method = "range", MARGIN = 2)

# make a distance matrix
aqhdis <- vegdist(aqhstand, method = "bray")

## plasma only
plasmah <- aqh %>% 
  filter(tissue.type == "plasma") %>% 
    mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))

# this data has an extra column; remove for standardization
plasmst <- plasmah %>% select(-c(id, Label, mouseID, treatmentID, Exercise, Weight, tissue.type, Time))

# standardize the data
plasmahstand <- decostand(plasmst, method = "range", MARGIN = 2)

# make a distance matrix
plasmahdis <- vegdist(plasmahstand, method = "bray")

## tumor only
tumorh <- aqh %>% 
  filter(tissue.type == "tumor")

# standardize the data
tumorhstand <- decostand(tumorh[ , 8:ncol(tumorh)], method = "range", MARGIN = 2)

# make a distance matrix
tumorhdis <- vegdist(tumorhstand, method = "bray")

```


## Plasma vs Tumor

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(aqhdis ~ tissue.type, data = aqh)

```


## Plasma D7 vs D21 vs D35

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(plasmahdis ~ Time, data = plasmah)

```


## Tumor


### Four Treatment Groups (2x2 factorial)

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(tumorhdis ~ treatmentID, data = tumorh)

```

### Exercise vs. Sedentary


```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(tumorhdis ~ Exercise, data = tumorh)

```

### Weight gain vs. weight maintenance

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(tumorhdis ~ Weight, data = tumorh)

```

## Plasma


### Exercise vs. Sedentary

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(plasmahdis ~ Exercise, data = plasmah)

```

### Weight gain vs. weight maintenance

```{r}
# perform pairwise ADONIS for the treatment group

pairwise.adonis2(plasmahdis ~ Weight, data = plasmah)

```