---
title: "Metabolomics Heatmaps"
author: "Emily Bean"
date: "March 25, 2020"
output: 
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE,
                      # figure margins
                      fig.margin = TRUE,
                      fig.fullwidth = TRUE,
                      fig.align = "center")

```

# Overview

This script makes heatmaps for the metabolites that were found to have significant changes (p < 0.05) in the `metabolomicsAnalysis.md` script.

If many metabolites have significant changes, the first 25 are shown on the heatmap (sorted by F statistic).

```{r}

## Read in all data and load packages
require(dplyr)
require(tidyr)
#require(vegan)
require(ggplot2)
require(emmeans)
require(stringr)
require(tibble)
require(gplots)
# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

# make separate tumor and plasma dataframes for easier analysis 
plasma <- aq %>% 
  filter(!Label == "tumor") %>% 
  mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))
# tumor df 
tumor <- filter(aq, Label == "tumor")

```

```{r}
## define function to perform an ANOVA, post-hoc emmeans, and return significant data

myAnova <- function(variable, df, metabList) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(metabList)) {
    
    # create function
    f <- as.formula(paste0("area ~", variable))
    
    # perform ANOVA
    mod <- aov(f, data = filter(df, metabolite == metabList[i]))
    
    # collect variables
    pvals <- data.frame(metabolite = metabList[i],
                        fStat =  round(summary(mod)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e <- as.data.frame(emmeans(mod, f1)$contrasts,
                       rownames = FALSE)
    
    # append to outDF
    out <- cbind(pvals, e)
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}
```

```{r}
# define a function to perform the ANOVA and post-hoc for each of the three Times


myAnovaTime <- function(variable, df, metabList) {
  
  # make empty dataframe
  sigsdf <- data.frame()
  
  # perform ANOVA and emmeans post-hoc
  for(i in 1:length(metabList)) {
    
    # create function
    f <- as.formula(paste0("area ~", variable))
    
    # perform ANOVA
    mod1 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "7"))
    
    # collect variables
    pvals1 <- data.frame(metabolite = metabList[i],
                         Time = "D7",
                        fStat =  round(summary(mod1)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    f1 <- as.formula(paste0("pairwise ~", variable))
    e1 <- as.data.frame(emmeans(mod1, f1)$contrasts,
                       rownames = FALSE)
    
    ## ANOVA Day 21
    # perform ANOVA
    mod2 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "21"))
    
    # collect variables
    pvals2 <- data.frame(metabolite = metabList[i],
                         Time = "D21",
                         fStat =  round(summary(mod2)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e2 <- as.data.frame(emmeans(mod2, f1)$contrasts,
                        rownames = FALSE)
    
    ## ANOVA Day 35
    # perform ANOVA
    mod3 <- aov(f, data = filter(df, metabolite == metabList[i] & Time == "35"))
    
    # collect variables
    pvals3 <- data.frame(metabolite = metabList[i],
                         Time = "D35",
                         fStat =  round(summary(mod3)[[1]][["F value"]][1], 4))
    
    # perform emmeans post-hoc
    e3 <- as.data.frame(emmeans(mod3, f1)$contrasts,
                        rownames = FALSE)
    
    # append to outDF
    out <- rbind(cbind(pvals1, e1),
                  cbind(pvals2, e2),
                  cbind(pvals3, e3))
    
    # get only significant p values to send out
    out <- out[out$p.value < 0.05, ] 
    
    # return dataframe of significant variables
    sigsdf <- rbind(sigsdf, out)
    
    
  }
  
  # round variables to make it easier to read
  sigsdf$SE <- round(sigsdf$SE, 3)
  sigsdf$t.ratio <- round(sigsdf$t.ratio, 3)
  sigsdf$p.value <- round(sigsdf$p.value, 3)
  
  # return the dataframe of significant variables
  
  return(sigsdf)
}

```


## Plasma v Tumor

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "tissue.type",
                   df = aq,
                   metabList = unique(aq$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  head(25)

# get significant metabolites
sigs <- aq[aq$metabolite %in% mod$metabolite, ] 

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, tissue.type, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

## Plasma D7 v D21 v D35


```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Time",
                   df = plasma,
                   metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))
  

# get significant metabolites
sigs <- plasma[plasma$metabolite %in% mod$metabolite, ] 

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Time, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

## Tumor

### 4 treatment groups

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "treatmentID",
                   df = tumor,
                   metabList = unique(tumor$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))
  

# get significant metabolites
sigs <- tumor[tumor$metabolite %in% mod$metabolite, ] 

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, treatmentID, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

### Exercise vs Sedentary

There is only 1 significant metabolite (Quinolinate) so we can't draw a heatmap.

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Exercise",
                   df = tumor,
                   metabList = unique(tumor$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))
  


```


### Weight gain vs maintenance

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Weight",
                   df = tumor,
                   metabList = unique(tumor$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))
  

# get significant metabolites
sigs <- tumor[tumor$metabolite %in% mod$metabolite, ] 

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Weight, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

## Plasma

### 4 treatment groups

#### D7

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "treatmentID",
               df = filter(plasma, Time == "7"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "7" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, treatmentID, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

#### D21

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "treatmentID",
               df = filter(plasma, Time == "21"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "21" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, treatmentID, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

#### D35

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "treatmentID",
               df = filter(plasma, Time == "35"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "35" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, treatmentID, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```


### Exercise vs Sedentary


#### D7

This also only has 1 significant metabolite.


#### D21


```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Exercise",
               df = filter(plasma, Time == "21"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "21" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Exercise, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```


#### D35

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Exercise",
               df = filter(plasma, Time == "35"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "35" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Exercise, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```


### Weight gain vs maintenance

#### D7

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Weight",
               df = filter(plasma, Time == "7"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "7" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Weight, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```


#### D21

```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Weight",
               df = filter(plasma, Time == "21"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "21" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Weight, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```

#### D35


```{r}
# ANOVA - get top 25 significant variables
mod <- myAnova(variable = "Weight",
               df = filter(plasma, Time == "35"),
               metabList = unique(plasma$metabolite)) %>%
  arrange(desc(fStat)) %>% 
  group_by(metabolite) %>% 
  top_n(n = 25, wt = desc(fStat))


# get significant metabolites
sigs <- plasma %>% 
  filter(Time == "35" & metabolite %in% mod$metabolite)

# parse into matrix for heatmap
sigsmat <- as.matrix(sigs %>% 
  # make unique identifying column
  mutate(idstr = str_remove_all(id, "160711-"),
         idcol = paste(idstr, Weight, sep = "_")) %>% 
  # make horizontal
  pivot_wider(names_from = metabolite, values_from = area) %>% 
  # make the identifying column into rownames
  column_to_rownames(var = "idcol") %>% 
    # keep only metabolites
  select_at(.vars = as.character(unique(sigs$metabolite)))
)

```


```{r}

# draw heatmap
heatmap.2(t(sigsmat), scale = "row", col = bluered(100), trace = "none", density.info = "none", margins = c(10, 12), cexRow = 1, cexCol = 1)

```