---
title: "Metabolomics PCA"
author: "Emily Bean"
date: "2/12/2020"
output:
  rmarkdown::github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, collapse = TRUE)

```


# Overview

This script follows the same structure as `metabolmicsAnalysis.Rmd` but performs a principal components analysis (PCA) on each data subset and then creates several diagnostics plots, then a grouping PCA and a biplot. 

The top eigenvalues are shown for each PCA.


# Pairwise comparisons  
*All comparisons made for both tumor and plasma tissues*

1. Plasma vs tumor
2. Plasma D7 vs D21 vd D35  
3. 4 treatment groups (2x2 factorial)  
4. Exercise vs sedentary  
5. Weight gain vs weight maintenance  



```{r}
## get data

require(dplyr)
require(tidyr)
require(factoextra)
require(FactoMineR)
require(corrplot)

# set ggplot2 them
ggplot2::theme_set(theme_bw())

# read custom PCA function
source("https://raw.githubusercontent.com/EmilyB17/ganda-lab/master/MyPCAFunction.R")

# read all aqueous data
aq <- read.table("https://raw.githubusercontent.com/EmilyB17/mice-metab/master/data/allAqueousCleaned.txt", header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(tissue.type =
           case_when(Label %in% "tumor" ~ "tumor",
                     Label %in% "plasmaD7" ~ "plasma",
                     Label %in% "plasmaD21" ~ "plasma",
                     Label %in% "plasmaD35" ~ "plasma"))

## Make all data horizontal for PCAs
aqh <- aq %>% 
  spread(key = metabolite, value = area)

# plasma only
plasmah <- aqh %>% 
  filter(tissue.type == "plasma") %>% 
    mutate(Time = factor(case_when(
    Label %in% "plasmaD7" ~ "7",
    Label %in% "plasmaD21" ~ "21",
    Label %in% "plasmaD35" ~ "35"
  ), ordered = TRUE, levels = c("7", "21", "35")))

# tumor only
tumorh <- aqh %>% 
  filter(tissue.type == "tumor")
```

## Plasma vs Tumor

The top 4 dimensions only explain 69% of the variance. The metabolites that seem to contribute most to the first 2 dimensions are N-acetyl-glutamine, Ribose, Citrulline, Indole-3-carboxylic acid, and Anthrilate.

Plasma and tumor tissue cluster very strongly; Glucose-1-6-phosphate and UDP D-glucuronate are both more correlated with the tumor tissue, while the other top metabolites are clustered with the plasma samples.

```{r}

# perform PCA
data <- aqh %>% select(-c(id, Label, mouseID, treatmentID, Exercise, Weight, tissue.type))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = aqh$tissue.type,
         groupname = "Tissue Type")
```

## Plasma D7 vs D21 vs D35

This data fits the PCA poorly; the first 6 dimensions only explain about 64% of the data (although this could be partially due to the large number of variables). However, the scree plot shows a large "elbow" between Dim2 and Dim3, so looking at the first two dimensions is sufficient. The top contributing metabolites are Citrulline, Trehalose-6-phosphate, 2-keto-D-gluconate, Hydroxyphenylacetic acid, and Threonine Homoserine.

The Time groups cluster all together, indicating that there is not a clear distinction between metabolic communities over time.

```{r}

# perform PCA
data <- plasmah %>% select(-c(id, Label, mouseID, treatmentID, Exercise, Weight, tissue.type, Time))

# print plots
myPCA(dat = data)
```


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$Time,
         groupname = "Time")
```

## Tumor


The scree plot shows a sharp decline between Dim1 and Dim2, as well as Dim2 and Dim3. The top contributing metabolites are Lactate, Phenylalanine, UDP-D-glucose, Ketoleucine, and Glutathione disulfide. The correlation plots show a clear distinction between groups of metabolites that cluster together.

The diagnostic plots show the PCA performed on the entire tumor dataset; there is no grouping variable involved.

```{r}

# perform PCA
data <- tumorh %>% select(-c(id, Label, mouseID, treatmentID, Exercise, Weight, tissue.type))

# print plots
myPCA(dat = data)
```


### Four Treatment Groups (2x2 factorial)

The 95% confidence intervals are very wide. Additionally, the correlating metabolites have short arrows, indicating a weak correlation. However, it seems that EX_ER and SED_ER are more tightly clustered than EX_AL and SED_AL.


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$treatmentID,
         groupname = "Treatment - Tumor")
```

### Exercise vs. Sedentary

The data is clustered together in a line with no apparent pattern.

```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$Exercise,
         groupname = "Exercise - Tumor")
```

### Weight gain vs. weight maintenance

In this group there is a tighter clustering in the ER group compared to the AL group. 
```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = tumorh$Weight,
         groupname = "Weight - Tumor")
```

## Plasma


There is also a fairly low percentage of variance explained by the top 6 eigenvalues (64% cumulative). The scree plot shows a sharp elbow between Dim2 and Dim3, so there is still a higher amount of variance explained by the first 2 axes. The top contributing metabolites are Citrulline, Trehalose-6-phosphate, 2-keto-D-gluconate, Hydroxyphenylacetic acid, and Threonine homoserine. Similarly to the tumor tissue, the plasma correlation plots seem to strongly cluster.


```{r}

# perform PCA
data <- plasmah %>% select(-c(id, Label, mouseID, treatmentID, Exercise, Weight, tissue.type, Time))

# print plots
myPCA(dat = data)
```

### Four Treatment Groups (2x2 factorial)

The arrows on the biplot are longer than the arrows in the tissue tumor, indicating a stronger correlation between the clusters and the top contributing metabolites. There is no clear pattern to the data.


```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$treatmentID,
         groupname = "Treatment - Plasma")
```

### Exercise vs. Sedentary

The first PCA plot shows that there are some samples with wide variance, notably most are in the Exercise group.

```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$Exercise,
         groupname = "Exercise - Plasma")
```

### Weight gain vs. weight maintenance

There is also no clear pattern to this data.

```{r}
# get grouped plots
plotPCAs(dat = data,
         grouping = plasmah$Weight,
         groupname = "Weight- Plasma")
```
